---
title: "Class3"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(xts)
```

## Time series analysis class 2

Notation:
$$
y: y_1, y_2,\ldots,y_T
$$

$$
y_{t - 1}: \text{ first lag}
$$
$$
y_t = y_{t - 1} + u_{t}
$$
## Purely random process (white noise)

$$
u_{t} \sim N(0, \sigma^2), \quad t=1,\ldots,T
$$
$$
E(u_t) = 0\\
Var(u_{t}) = \sigma^2
$$

Generate 100 values from a standard normal distribution (i.e. $\sigma^2  = 1$).

```{r}
randomValues <- rnorm(100)
timeIndex <- seq.Date(as.Date("2018-10-10"), by = "day", length.out = 100)
u <- xts(randomValues, timeIndex)
plot(u)
mean(u)
```


$$
X \sim N(\mu, \sigma^2)\\
\text{ roughly } 95\% \text{ of the realisations of X are expected to be in the interval} \\
[\mu - 2\sigma, \mu + 2\sigma] 
$$
$$
X \sim N(0, 1)\\
[ - 2, 2]
$$


```{r}
uL1 <- lag(u)
combinedSeries <- cbind(u, uL1)
plot(as.data.frame(uL1)[, 1], as.data.frame(u)[, 1], xlab = "u_{t - 1}", ylab = "u_{t}", main = "Scatterplot of u_t and u_{t - 1}")
```


```{r}
hist(as.data.frame(u)[, 1], main = "Histogram of u")
```

## Sample (empirical) covariance

$$
x: x_{1}, x_{2}, \ldots, x_{T}\\
y: y_{1}, y_{2}, \ldots, y_{T}\\
$$

$$
u_{t} \sim N(0, 1) \text{ Standard normal distribution}\\
y_{t} = x_{t} + 7 u_{t}
$$
## Expected value of $y_t$

$$
E(y_t) = E(x_t + 7u_t) = E(x_t) + E(7u_t) = x_{t} + E(7u_t) = x_{t} + 7\underbrace{E(u_t)}_{0} = x_t \\
E(y_t) = x_t \\
Var(7u_t) = 7^2 \underbrace{Var(u_t)}_{1} = 49
$$

$$
E(y_t) = \underbrace{E(x_t)}_{x_t} + \underbrace{E(7u_t)}_{0} \\
E(u_t) = 0 \text{ by the definition of the normal distribution with mean 0} \\
E(7u_t) = 7E(u_t) = 0 \text{ because 7 is a constant (non-random)}\\
Var(7u_t) = 7^2Var(u_t) = 7^2\times 1
$$



```{r}
x <- 1:50
noise <- 20*rnorm(50)
```

```{r}
y <- 100 * x + noise
```

$$
E(y_t) = E(x_t + u_t) = E(x_{t}) + \underbrace{E(u_{t})}_{0} = E(x_t)
$$

```{r}
plot(x, y)
mean(x)
mean(y)
abline(v = mean(x), col = 3, lwd = 2)
abline(h = mean(y), col = 2, lwd = 2)
```
```{r}
cov(x, y)
```



$$
(x_t - \overline{x})(y_{t} - \overline{y}) > 0 \text{ for points in the top right part}\\
(x_t - \overline{x})(y_{t} - \overline{y}) > 0 \text{ for points in the bottom left part}\\
(x_t - \overline{x})(y_{t} - \overline{y}) < 0 \text{ for points in the bottom right and top left parts}
$$

$$
Cov(x, y) = \frac{1}{T - 1}\sum_{t = 1}^{T}(x_t - \overline{x})(y_{t} - \overline{y})
$$


$$
\overline{y} = \frac{1}{T}\sum_{t = 1}^{T}y_{t} \\
(y_t - \overline{y}) > 0 \text{ for points above the red line}\\
(y_t - \overline{y}) < 0 \text{ for points below the red line}\\
(x_t - \overline{x}) > 0 \text{ for points to the right of the green line}\\
(x_t - \overline{x}) < 0 \text{ for points to the left of the green line}
$$

$$
s^2(x) = \frac{1}{T - 1}\sum_{t = 1}^{T}(x_t - \overline{x})^2\\
s^2(y) = \frac{1}{T - 1}\sum_{t = 1}^{T}(y_t - \overline{y})^2\\

\rho(x, y) = \frac{Cov(x, y)}{\sqrt{s^2(x)s^2(y)}} \text{ correlation coefficient}\\
-1 \leq \rho(x, y) \leq 1
$$

$$
y_{t} =  - x_{t}
$$

$$
\rho(x, y) = \rho(x, x) = \frac{Cov(x, x)}{\sqrt{s^2(x)s^2(x)}} = \frac{s^2(x)}{s^2(x)} = 1
$$

$$
Cov(x, x) = \frac{1}{T - 1}\sum_{t = 1}^{T}(x_t - \overline{x})^2 = Var(x) = s^2(x)
$$
$$
y_t = bx_{t} + a, \quad b,a \in R\\
b > 0: \rho(x, y) = 1\\
b < 0: \rho(x, y) = -1 \\
$$

$$
y_t, y_{t - 1}, y_{t - 2},\ldots, y_{t - k}
$$

$$
\gamma_0 = Cov(y_t, y_{t}) = Var(y_t) = s^2(x)\\
\gamma_{1} = Cov(y_{t}, y_{t - 1}) \text{ first order autocovariance }\\
\gamma_{2} = Cov(y_{t}, y_{t - 2}) \text{ second order autocovariance } \\
\gamma_{3} =Cov(y_{t}, y_{t - 3}) \text{ third order autocovariance } \\
\vdots \\
\gamma_{k} =Cov(y_{t}, y_{t - k}) \text{ k-th order autocovariance } \\
$$
$$
\rho_{0}= \rho(y_t, y_{t}) = 1 \\
\rho_{1} = \rho(y_{t}, y_{t - 1}) \text{ first order autocorrelation }\\
\rho_{2} = \rho(y_{t}, y_{t - 2}) \text{ second order autocorrelation } \\
\rho_{3} =\rho(y_{t}, y_{t - 3}) \text{ third order autocorrelation } \\
\vdots \\
\rho_{k} =\rho(y_{t}, y_{t - k}) \text{ k-th order autocorrelation } \\
$$

$$
\rho_1 = \frac{\gamma_1}{\gamma_0} \\
\rho_2 = \frac{\gamma_2}{\gamma_0} \\
\vdots\\
\rho_k = \frac{\gamma_k}{\gamma_0}\\
$$
$$
\text{Assume that } s^2(y_t) = s^2(y_{t - 1}) \\
\rho_{1} = \rho(y_t, y_{t - 1}) = \frac{Cov(y_t, y_{t - 1})}{\sqrt{s^2(y_t)s^2(y_{t})}} = \frac{Cov(y_t, y_{t - 1})}{s^2(y_t)} = \frac{\gamma_1}{\gamma_0}
$$

$$
u_t \sim N(0, 1) \\
u_{t} \text{ and } u_{t - 1} \text{ are not correlated}
$$

```{r}
library(xts)
n <- 100
uRaw <- rnorm(n) ## Generate 100 observations from a standard normal distribution
u <-xts(uRaw, order.by = seq.Date(as.Date("2020-11-09"), by = "day", length.out = n)) 
acf(u)
acf(u, plot = FALSE)
```
## Exercise

$$
x: x_1, x_2, x_3\\
x: (1, 3, 5)\\

\overline{x} = 9 / 3 = 3\\
\text{ Sample variance: } \gamma_0 = \frac{1}{3 - 1}[(1 - 3)^2 + (3 - 3)^2 + (5 - 3)^2] = \frac{1}{2}(2^2 + 0 +2^2 ) = \frac{1}{2}2*4 = 4 \\
\text{ Sample covariance: }\gamma_1 = ? \\
\text{ Sample correlation: } \rho_1 = ?
$$
$$
\gamma_1 = Cov(x_t, x_{t - 1}) = \frac{1}{T - 1}\sum_{t = 1}^{T}(x_{t} - \overline{x})(x_{t - 1} - \overline{x})
$$
$$
\gamma_1 = \frac{1}{3 - 1}[(3 - 3)(1 - 3) + (5 - 3)(3 - 3)] = 0\\
\rho_1 = \frac{\gamma_1}{\gamma_0} = \frac{0}{4} = 0
$$

$$
s^2(x) = \frac{1}{T - 1}\sum_{t = 1}^{T}(x_t - \overline{x})^2\\
\rho(x, y) = \frac{Cov(x, y)}{\sqrt{s^2(x)s^2(y)}} \text{ correlation coefficient}\\
-1 \leq \rho(x, y) \leq 1
$$

$$
\rho(x, y) = \rho(x, x) = \frac{Cov(x, x)}{\sqrt{s^2(x)s^2(x)}} = \frac{s^2(x)}{s^2(x)} = 1
$$

```{r}
x <- c(1, 3, 5)
var(x)
```

```{r}
n <- 100
uRaw <- 100 * runif(n) ## Generate 100 observations from a standard normal distribution
u <-xts(uRaw, order.by = seq.Date(as.Date("2020-11-09"), by = "day", length.out = n)) 
uL1 <- lag(u)
plot(as.data.frame(uL1)[, 1], as.data.frame(u)[, 1])
abline(h = 0)
abline(v = 0)
```

```{r}
plot(u)
```

## Structure of the data (realm of the data)
- (sample) Mean value of the series (calculated from actual data)
- (sample) Variance of the series (calculated from actual data)
- (sample) Autocorrelation structure (calculated from actual data)
$$
\rho_{1}, \rho_{2}, \ldots, \rho_{k}
$$

## Structure of the model (realm of mathematical models)

- Expected value of the series (model)
- Variance of the series (model)
- Autocorrelation structure (model)



```{r}
plot(u)
```
```{r}
uL1 <- lag(u)
plot(as.data.frame(uL1)[, 1], as.data.frame(u)[, 1])
```



$$
y_{1}, y_{2}\ldots,y_{T}\\
y_{0}, y_{1}\ldots,y_{T - 1} : \text{first lag of } y_{t}: y_{t - 1}
$$

## Time series structure

- level / trend of the time series
- fluctuations (variance)
- auto-covariance / auto-correlation structure

$$
y_{t}\\
\rho_{1} = Corr(y_{t}, y_{t - 1}) \text{ first order autocorrelation coefficient}\\
\rho_{2} = Corr(y_{t}, y_{t - 2}) \text{ second order autocorrelation coefficient}\\
\rho_{\tau} = Corr(y_{t}, y_{t - \tau})
$$

$$
\rho_{0} = Corr(y_{t}, y_{t})
$$


```{r}
randomValues <- rnorm(100)
timeIndex <- seq.Date(as.Date("2018-10-10"), by = "day", length.out = 100)
u <- xts(randomValues, timeIndex)
## plot(u)
## mean(u)
acf(u)
```

```{r}
## Auto-correlation function

```
$$
y_{t} = y_{t} \\
Corr(y_{t}, y_{t}) = 1
$$
## Blue dotted lines in the autocorrelation plot
- 95% confidence interval for the autocorrelations

$$

$$

$$
a,b \in R\\
b > 0\\
y_{t} = a + b x_{t}, \quad \rho_{x, y} = 1\\
y_{t} = a - b x_{t}, \quad \rho_{x, y} = -1
$$















